---
title: "HW1 - Tidycensus Markdown (Resubmit)"
author: "Jun, Youngsang"
date: "(First) September 5, 2024 (Resubmit) September 9, 2024"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

# 1. Rerun the Code

Install and load Libraries

```{r setup, include=FALSE, cache = TRUE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(rmarkdown)
```

```{r setup_packages1, warning = FALSE, eval = FALSE}
install.packages('tidyverse')
install.packages('tidycensus')
install.packages('sf')
install.packages('pander')
install.packages('kableextra')
```

```{r setup_packages2, warning = FALSE, message = FALSE}
library(tidyverse)
library(tidycensus)
library(sf)
library(pander)
library(kableExtra)
```

Census API Key

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("b2835f54d89a4499ba29829c908967b86765b345", overwrite = TRUE)
```

Load census data dictionaries

```{r load_variables, cache = TRUE}

acs_variable_list.2020 <- load_variables(2020, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)

acs_variable_list.2016 <- load_variables(2016, #year
                                         "acs5", #five year ACS estimates
                                         cache = TRUE)
```

Create a vector of census variables

```{r acs_vars}
acs_vars <- c("B01001_001E", # ACS total Pop estimate
              "B25002_001E", # Estimate of total housing units
              "B25002_003E", # Number of vacant housing units
              "B19013_001E", # Median HH Income ($)
              "B02001_002E", # People describing themselves as "white alone"
              "B06009_006E") # Total graduate or professional degree
```

Call the Census API to get tract level data for 2020 for all of Philadelphia

```{r get_acs_2020, cache = TRUE, message = FALSE, warning = FALSE}
acsTractsPHL.2020 <- get_acs(geography = "tract",
                             year = 2020, 
                             variables = acs_vars, 
                             geometry = FALSE, 
                             state = "PA", 
                             county = "Philadelphia", 
                             output = "wide") 
```

Mutating, selecting and renaming variables

```{r do_some_dplyr, cache = TRUE}
acsTractsPHL.2020 <- acsTractsPHL.2020 %>%
  dplyr::select (GEOID, NAME, all_of(acs_vars))

acsTractsPHL.2020 <- acsTractsPHL.2020 %>%
  rename (total_pop.2020 = B01001_001E,
          total_HU.2020 = B25002_001E,
          total_vacant.2020 = B25002_003E,
          med_HH_Income.2020 = B19013_001E,
          total_White.2020 = B02001_002E,
          total_GradDeg.2020 = B06009_006E)

acsTractsPHL.2020 <- acsTractsPHL.2020 %>%
  mutate(vacancyPct.2020 = total_vacant.2020/total_HU.2020,
         pctWhite.2020   = total_White.2020/total_pop.2020)
```

```{r get_acs_2016, cache = TRUE, message = FALSE}
acsTractsPHL.2016 <- get_acs(geography = "tract",
                             year = 2016, 
                             variables = acs_vars,
                             geometry = FALSE,
                             state = "PA", 
                             county = "Philadelphia",
                             output = "wide") %>%
  dplyr::select (GEOID, NAME, all_of(acs_vars)) %>% 
  rename (total_pop.2016 = B01001_001E,
          total_HU.2016 = B25002_001E,
          total_vacant.2016 = B25002_003E,
          med_HH_Income.2016 = B19013_001E,
          total_White.2016 = B02001_002E,
          total_GradDeg.2016 = B06009_006E) %>%
  mutate(vacancyPct.2016 = total_vacant.2016/total_HU.2016,
         pctWhite.2016 = total_White.2016/total_pop.2016)
```

Joining data

```{r left_join_tracts, cache = TRUE}
allACS <- left_join(acsTractsPHL.2016, acsTractsPHL.2020,
                    by= c("GEOID"))
```

Doing column math using mutate

```{r do_mutates, cache = TRUE}
allACS <- allACS %>%
  mutate(change_med_HH_Income = med_HH_Income.2020 - (med_HH_Income.2016 * 1.08), 
         change_Grad_Degree_Pct = (total_GradDeg.2020/total_pop.2020)-(total_GradDeg.2016/total_pop.2016))

```

Exploring central tendencies

```{r base_r_summaries_NA_example, cache = TRUE}
mean(allACS$change_med_HH_Income)
```

```{r base_r_summaries_NA_rm, cache = TRUE}
mean(allACS$change_med_HH_Income, na.rm = TRUE)

median(allACS$change_med_HH_Income, na.rm = TRUE)
```

Exploring distributions

```{r basic_histogram, cache = TRUE}
hist(allACS$change_med_HH_Income)
```

```{r ggplot_histogram_simple, warning = FALSE, cache = TRUE, message = FALSE}
ggplot(allACS) +
  geom_histogram(aes(change_med_HH_Income)) +
  theme_minimal()
```

```{r ggplot_histogram, warning=FALSE, cache=FALSE}
ggplot(allACS)+
  geom_histogram(aes(change_med_HH_Income), binwidth = 5000)+
  labs(
    title = "Change in Philadelphia HH median income by tract, 2016-2020",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x="Change in Med HH Income (2020 dollars)", 
       y="Number of tracts") +
  theme_minimal()
```

Making a summary table

```{r summaryTable, cache = TRUE}


summaryTable <- allACS %>%
  summarize(mean_change_HH_Income = mean(change_med_HH_Income, na.rm = TRUE),
            med_change_HH_Income = median(change_med_HH_Income, na.rm = TRUE))

pander(summaryTable)
```

Comparing geographies

```{r myTracts, cache = TRUE}

myTracts <- c("42101023500", 
              "42101023600", 
              "42101023700", 
              "42101025300", 
              "42101025400",
              "42101025500", 
              "42101025600", 
              "42101038800")

allACS <- allACS %>%
  mutate(mtAiry = ifelse(GEOID %in% myTracts, "MT AIRY", "REST OF PHILADELPHIA"))
```

```{r summary_table_2, cache = TRUE, message = FALSE, warning = FALSE}
summaryTable2 <- allACS %>%
  group_by(mtAiry) %>%
  summarize(mean_change_HH_Income = mean(change_med_HH_Income, na.rm = TRUE),
            med_change_HH_Income = median(change_med_HH_Income, na.rm = TRUE))

pander(summaryTable2)
```

Graphic comparisons Using ggplot2

```{r ggplot_histogram_2, warning = FALSE, cache = TRUE}
ggplot(allACS)+
  geom_histogram(aes(change_med_HH_Income),
                 binwidth = 5000)+
  labs(
    title = "Change in Philadelphia HH median income by tract, 2016-2020",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x="Change in Med HH Income (2020 dollars)", 
       y="Number of tracts")+
  facet_wrap(~mtAiry, scales = "free") + 
  theme_minimal()
  
```

```{r ggplot_point1, warning = FALSE, cache = TRUE}
ggplot(allACS)+
  geom_point(aes(x =med_HH_Income.2016 * 1.08, 
                 y = med_HH_Income.2020,
                 color = mtAiry))+
  geom_abline(intercept = 0, slope = 1)+
  labs(
    title = "2020 Median HH Income as a Function of 2016 Median HH Income",
    subtitle = "All figures in 2020 dollars",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x="Med HH Income 2016 ($)", 
    y="Med HH Income 2020 ($)") +
  theme_minimal()
```

```{r ggplot_point2, warning = FALSE, cache = TRUE, message = FALSE}
ggplot(allACS)+
  geom_point(aes(x = 100* pctWhite.2020, 
                 y = med_HH_Income.2020,
                 color = mtAiry))+
  geom_smooth(aes(x = 100* pctWhite.2020, 
                  y = med_HH_Income.2020), 
              method = "lm", se = FALSE)+
  labs(
    title = "2020 Median HH Income as a Function of Pct White",
    subtitle = "All figures in 2020 dollars",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x="Pct. Residents Identifying as 'White Only'", 
    y="Med HH Income 2020 ($)") +
  theme_minimal()
```

# 2. Create a Code Chunks

## a. A ggplot plot

Plot number of vacant housing units per tract in 2020 as a function of vacant housing units in 2016, (e.g. the change in raw vacant housing units from 2016-2020 - using "B25002_003E", \# Number of vacant housing units, part of the `acs_vars` vector). Use the ggplot "color" parameter in the geom_point aesthetics to visualize which are the Mt. Airy tracts. Annotate this plot with a single sentence describing what it shows (not as a plot title, but within the text of the markdown).

```{r ggplot_point3, warning = FALSE, cache = TRUE}
ggplot(allACS)+
  geom_point(aes(x = total_vacant.2016, 
                 y = total_vacant.2020,
                 color = mtAiry))+
  geom_abline(intercept = 0, slope = 1)+
  scale_x_continuous(breaks = seq(0, 1000, by = 200)) +
  scale_y_continuous(breaks = seq(0, 1000, by = 200)) +
  labs(
    title = "Vacant Housing Units in 2020 as a Function of Vacant Housing Units in 2016",
    subtitle = "Count by Tract",
    caption = "Data: US Census Bureau, ACS 5-year estimates",
    x="Vacant Housing 2016 (count)", 
    y="Vacant Housing 2020 (count)",
    color = "Legend") +
  theme_minimal()
```

### Single Sentence Explanation

Comparing 2016, three tracts of Mt. Airy's number of vacant housing units per tract increased in 2020, one tract decreased, and two had no change.

## b. A ggplot map

Map a ggplot map symbolizing the number of vacant housing units in 2016 as a percentage of total housing units (e.g. "total_HU.2016") on a per-tract basis. Outline the Mt. Airy neighborhood on the map. Annotate this map with a single sentence describing what it shows or implies about Mt Airy and Philadelphia.

```{r spatial_tidycensus2, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
acsTractsPHL.2016.sf <- get_acs(geography = "tract",
                             year = 2016, 
                             variables = acs_vars, 
                             geometry = TRUE, 
                             state = "PA", 
                             county = "Philadelphia", 
                             output = "wide") %>% 
  dplyr::select (GEOID, NAME, all_of(acs_vars)) %>%
  rename (total_pop.2016 = B01001_001E,
          total_HU.2016 = B25002_001E,
          total_vacant.2016 = B25002_003E,
          med_HH_Income.2016 = B19013_001E,
          total_White.2016 = B02001_002E,
          total_GradDeg.2016 = B06009_006E) %>%
  mutate(vacancyPct.2016 = total_vacant.2016/total_HU.2016) %>%
  mutate(mtAiry = ifelse(GEOID %in% myTracts, "MT AIRY", "REST OF PHILADELPHIA")) %>%
  st_as_sf(crs = 4326) # Turn shp into sf object and project as WGS84
```

```{r ggplot_geom_sf2, warning = FALSE, cache = TRUE}
ggplot()+
  geom_sf(data = acsTractsPHL.2016.sf, aes(fill = vacancyPct.2016),
          color = "gray")+
  scale_fill_gradient2(midpoint = 0.0) +
  geom_sf(data = acsTractsPHL.2016.sf %>%
            filter(mtAiry == "MT AIRY") %>%
            st_union(),
          color = "#FF5440",
          fill = "transparent")+
  labs(
    title = "Vacant Housing Units Rate out of Total Housing Units by Tract in 2016",
    caption = "Data: US Census Bureau, ACS 5-year estimates") +
  theme_void()
```

### Single Sentence Explanation

According to the map, the vacancy rate of the Mt. Airy area in 2016 was lower than the north-central and southwest of Philadelphia County, and some tracts southeast of Mt. Airy are higher than the north of Mt. Airy area.

## c. Summary Table

A code chunk that renders a "kable" table showing a summary of mean number of vacant housing units per tract for Mt. Airy vs. the rest Philadelphia as a whole in 2020. Annotate this plot with a single sentence describing what it shows or implies about the data. Install the kable and kableExtra packages to do this - there is a simple tutorial on that here:

```{r summary_table_3, cache = TRUE, message = FALSE, warning = FALSE}
summaryTable3 <- allACS %>%
  group_by(mtAiry) %>%
  summarize(mean_total_vacant_2020 = mean(total_vacant.2020, na.rm = TRUE))

summaryTable3 %>%
  kbl(caption = "Mean Number of Vacant Housing Units per Tract in 2020",
      col.names = c("AREA", "MEAN")) %>%
  kable_styling(position="center") %>%
  kable_classic(full_width = T, html_font = "Arial")
```

### Single Sentence Explanation

Mt. Airy's mean number of vacant housing units per tract is 156.625, lower than the rest of Philadelphia County's mean of 196.893.
